---
title: Using RcppProgress
output: rmarkdown::html_vignette
toc: true
vignette: >
  %\VignetteIndexEntry{Using RcppProgress}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

 Suppose that you have this `long_computation()` C++ function that you call from R using Rcpp:

```{r long_computation, engine="Rcpp"}
// [[Rcpp::depends(RcppProgress)]]
#include <progress.hpp>
// [[Rcpp::export]]
double long_computation(int nb) {
    double sum = 0;
    for (int i = 0; i < nb; ++i) {
        for (int j = 0; j < nb; ++j) {
	    sum += R::dlnorm(i+j, 0.0, 1.0, 0);
	}
    }
    return sum + nb;
}
```

```{r long_computation_call}
system.time(res  <- long_computation(1000))
print(res)
```


## Checking for user interrupts

Let's modify our code to add a check for user interruption by calling the function
`Progress::check_abort()`.  Note the `Rcpp::depends(RcppProgress)` attribute in
the header part that takes care of the include path for the *progress.hpp*
header.

Now the `long_computation_interruptible()` call is interruptible (e.g. by typing CTRL+C in the
classic R console).

```{r long_computation_interruptible, engine="Rcpp"}
// [[Rcpp::depends(RcppProgress)]]
#include <progress.hpp>
// [[Rcpp::export]]
double long_computation_interruptible(int nb) {
    double sum = 0;
    for (int i = 0; i < nb; ++i) {
        if (Progress::check_abort() )
            return -1.0;
  	for (int j = 0; j < nb; ++j) {
	    sum += R::dlnorm(i+j, 0.0, 1.0, 0);
	}
    }
    return sum + nb;
}
```

```{r long_computation_interruptible_call}
system.time(res  <- long_computation_interruptible(3000)) # interrupt me
print(res)
```

You may wonder why we put the `check_abort()` call in the first loop instead
that in the second.  The performance cost of this call is not negligible. 
It should be put in a place called often enough (like once per second) 
yet not too often to minimize the overhead.

## Adding a progress bar
  
Time to add the progress bar. The `Progress::increment()` function is quite fast, so we
can put it in the second loop.  In real life example, it is sufficient to put
it at a place called at least every second.
 
```{r long_computation_progress, engine="Rcpp"}
// [[Rcpp::depends(RcppProgress)]]
#include <progress.hpp>
#include <progress_bar.hpp>
// [[Rcpp::export]]
double long_computation_progress(int nb, bool display_progress=true) {
    double sum = 0;
    Progress p(nb*nb, display_progress);
    for (int i = 0; i < nb; ++i) {
        if (Progress::check_abort() )
            return -1.0;
        for (int j = 0; j < nb; ++j) {
            p.increment(); // update progress
	    sum += R::dlnorm(i+j, 0.0, 1.0, 0);
	}
    }
    return sum + nb;
}
```

```{r long_computation_progress_call}
system.time(res  <- long_computation_progress(3000)) # interrupt me
print(res)
```

## OpenMP support

First we need this to enable OpenMP support for `gcc`. In the early days we used

```
Sys.setenv("PKG_CXXFLAGS"="-fopenmp")
Sys.setenv("PKG_LIBS"="-fopenmp")
```

and more recent version of Rcpp have a plugin  which does this for us.

Here is an OpenMP version of our function:

```{r long_computation_omp, engine="Rcpp"}
#ifdef _OPENMP
#include <omp.h>
#endif
// [[Rcpp::plugins(openmp)]]
// [[Rcpp::depends(RcppProgress)]]
#include <progress.hpp>
// [[Rcpp::export]]
double long_computation_omp(int nb, int threads=1) {
#ifdef _OPENMP
    if ( threads > 0 )
        omp_set_num_threads( threads );
    REprintf("Number of threads=%i\n", omp_get_max_threads());
#endif
 
    double sum = 0;
#pragma omp parallel for schedule(dynamic)   
    for (int i = 0; i < nb; ++i) {
        double thread_sum = 0;
  	for (int j = 0; j < nb; ++j) {
	    thread_sum += R::dlnorm(i+j, 0.0, 1.0, 0);
	}
        sum += thread_sum;
    }
    return sum + nb;
}
```

Now check that it is parallelized. The execution time for the first call that uses 4 threads should
be much faster (~ 3 times faster on my computer) than the call with one single thread:

```{r long_computation_omp_call}
system.time(res4 <- long_computation_omp(5000, 4))
print(res4)
system.time(res1 <- long_computation_omp(5000, 1))
print(res1)
```

## adding progress monitoring to the openMP function

```{r long_computation_omp_progress, engine="Rcpp", eval=FALSE}
#ifdef _OPENMP
#include <omp.h>
#endif
// [[Rcpp::plugins(openmp)]]
// [[Rcpp::depends(RcppProgress)]]
#include <progress.hpp>
#include <progress_bar.hpp>
// [[Rcpp::export]]
double long_computation_omp_progress(const int nb, int threads=1) {
#ifdef _OPENMP
    if ( threads > 0 )
        omp_set_num_threads( threads );
#endif
    Progress p(nb, true);
    double sum = 0;
#pragma omp parallel for default(none) reduction(+ : sum) schedule(dynamic)
    for (int i = 0; i < nb; ++i) {
        double thread_sum = 0;
        if ( ! Progress::check_abort() ) {
            p.increment(); // update progress
            for (int j = 0; j < nb; ++j) {
                thread_sum += R::dlnorm(i+j, 0.0, 1.0, 0);
            }
        }
        sum += thread_sum;
    }
    return sum + nb;
}
```

```{r long_computation_omp_progress_call, eval=FALSE}
system.time(long_computation_omp_progress(5000, 4))
```

## Test it now

If you want to test it now in your R console, just paste the following code
(after installing **RcppProgress** of course):

```{r final_example, eval=TRUE}
CODE <- r"(
#ifdef _OPENMP
#include <omp.h>
#endif
// [[Rcpp::plugins(openmp)]]
// [[Rcpp::depends(RcppProgress)]]
#include <progress.hpp>
#include <progress_bar.hpp>

// [[Rcpp::export]]
double long_computation_omp_progress(int nb, int threads=1) {
#ifdef _OPENMP
    if ( threads > 0 )
        omp_set_num_threads( threads );
#endif
    Progress p(nb, true);
    double sum = 0;
#pragma omp parallel for schedule(dynamic)   
    for (int i = 0; i < nb; ++i) {
        double thread_sum = 0;
        if ( ! Progress::check_abort() ) {
            p.increment(); // update progress
            for (int j = 0; j < nb; ++j) {
                thread_sum += R::dlnorm(i+j, 0.0, 1.0, 0);
            }
        }
        sum += thread_sum;
    }
  
    return sum + nb;
}
)"
```

and run

```{r final_example_call}
Rcpp::sourceCpp(code = CODE)
res <- long_computation_omp_progress(10000, 4)
```

